---
title: "NovClusterprofiler184spl"
author: "Saul"
date: "2023-11-16"
output: html_document
---

```{r}
#save(DvB_uniqgenes, file = "DvB_uniqgenes.RData")

https://bioconductor.org/packages/release/bioc/vignettes/clusterProfiler/inst/doc/clusterProfiler.html

#Usaremos el df con los geneIDs de ENSEMBLE en rownames

BiocManager::install("clusterProfiler", version = 3.18)
BiocManager::install("pathview")
BiocManager::install("enrichplot")
library(clusterProfiler)
library(enrichplot)
# we use ggplot2 to add x axis labels (ex: ridgeplot)
library(ggplot2)

# SET THE DESIRED ORGANISM
https://bioconductor.org/packages/release/BiocViews.html#___OrgDb

organism = "org.Hs.eg.db"

#BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)

#Prepare Input

# data from deseq2
DvB_uniqgenes
SexFvM_uniqgenes
AgeOvY_uniqgenes
Cluster1_uniqgenes
Cluster2_uniqgenes
Cluster3_uniqgenes

# we want the log2 fold change 
original_gene_listDvB    <- DvB_uniqgenes$log2FoldChange
original_gene_listSexFvM <- SexFvM_uniqgenes$log2FoldChange
original_gene_listAgeOvY <- AgeOvY_uniqgenes$log2FoldChange
original_gene_listC1 <- Cluster1_uniqgenes$log2FoldChange
original_gene_listC2 <- Cluster2_uniqgenes$log2FoldChange
original_gene_listC3 <- Cluster3_uniqgenes$log2FoldChange

# name the vector
names(original_gene_listDvB) <- DvB_uniqgenes$external_gene_name
names(original_gene_listSexFvM) <- SexFvM_uniqgenes$external_gene_name
names(original_gene_listAgeOvY) <- AgeOvY_uniqgenes$external_gene_name
names(original_gene_listC1) <- Cluster1_uniqgenes$external_gene_name
names(original_gene_listC2) <- Cluster2_uniqgenes$external_gene_name
names(original_gene_listC3) <- Cluster3_uniqgenes$external_gene_name

# omit any NA values 
gene_listDvB<-na.omit(original_gene_listDvB)
gene_listSexFvM<-na.omit(original_gene_listSexFvM)
gene_listAgeOvY<-na.omit(original_gene_listAgeOvY)
gene_listC1<-na.omit(original_gene_listC1)
gene_listC2<-na.omit(original_gene_listC2)
gene_listC3<-na.omit(original_gene_listC3)


# sort the list in decreasing order (required for clusterProfiler)
gene_listDvB = sort(gene_listDvB, decreasing = TRUE)
gene_listSexFvM = sort(gene_listSexFvM, decreasing = TRUE)
gene_listAgeOvY = sort(gene_listAgeOvY, decreasing = TRUE)
gene_listC1 = sort(gene_listC1, decreasing = TRUE)
gene_listC2 = sort(gene_listC2, decreasing = TRUE)
gene_listC3 = sort(gene_listC3, decreasing = TRUE)

## Gene Set Enrichment
Params:  
  
**keyType** This is the source of the annotation (gene ids). The options vary for each annotation. In the example of *org.Dm.eg.db*, the options are:   
  
"ACCNUM"       "ALIAS"        "ENSEMBL"      "ENSEMBLPROT"  "ENSEMBLTRANS" "ENTREZID"      
"ENZYME"       "EVIDENCE"     "EVIDENCEALL"  "FLYBASE"      "FLYBASECG"    "FLYBASEPROT"   
"GENENAME"     "GO"           "GOALL"        "MAP"          "ONTOLOGY"     "ONTOLOGYALL"   
"PATH"         "PMID"         "REFSEQ"       "SYMBOL"       "UNIGENE"      "UNIPROT"  
  
Check which options are available with the `keytypes` command, for example `keytypes(org.Hs.eg.db)`. 

keytypes(org.Hs.eg.db)
  
**ont** one of "BP", "MF", "CC" or "ALL"  
**nPerm** permutation numbers, the higher the number of permutations you set, the more accurate your results is, but it will also cost longer time for running permutation.  
**minGSSize** minimal size of each geneSet for analyzing.   
**maxGSSize** maximal size of genes annotated for testing.   
**pvalueCutoff** pvalue Cutoff.   
**pAdjustMethod** one of "holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", "none" 

gseDvB <-    gseGO(gene_listDvB, ont = "ALL", keyType = "SYMBOL", OrgDb = "org.Hs.eg.db")
gseSexFvM <- gseGO(gene_listSexFvM, ont = "ALL", keyType = "SYMBOL", OrgDb = "org.Hs.eg.db")
gseAgeOvY <- gseGO(gene_listAgeOvY, ont = "ALL", keyType = "SYMBOL", OrgDb = "org.Hs.eg.db")
gseC1 <-     gseGO(gene_listC1, ont = "ALL", keyType = "SYMBOL", OrgDb = "org.Hs.eg.db")
gseC2 <-     gseGO(gene_listC2, ont = "ALL", keyType = "SYMBOL", OrgDb = "org.Hs.eg.db")
gseC3 <-     gseGO(gene_listC3, ont = "ALL", keyType = "SYMBOL", OrgDb = "org.Hs.eg.db")

save(gseDvB, file =   "gseDvB.RData")
save(gseSexFvM, file ="gseSexFvM.RData")
save(gseAgeOvY, file ="gseAgeOvY.RData")
save(gseC1, file =    "gseC1.RData")
save(gseC2, file =    "gseC2.RData")
save(gseC3, file =    "gseC3.RData")

```
##Dotplot
```{r echo=TRUE, fig.width=15, fig.height=8}
require(DOSE)
dotplot(gseDvB, showCategory=10, split=".sign") + facet_grid(.~.sign)
dotplot(gseSexFvM, showCategory=10, split=".sign") + facet_grid(.~.sign)
dotplot(gseAgeOvY, showCategory=10, split=".sign") + facet_grid(.~.sign)
dotplot(gseC1, showCategory=10, split=".sign") + facet_grid(.~.sign)
dotplot(gseC2, showCategory=10, split=".sign") + facet_grid(.~.sign)
dotplot(gseC3, showCategory=10, split=".sign") + facet_grid(.~.sign)

```

##Encrichment plot map:
Enrichment map organizes enriched terms into a network with edges connecting overlapping gene sets. In this way, mutually overlapping gene sets are tend to cluster together, making it easy to identify functional modules.
```{r echo=TRUE}
#library(GOSemSim)
emapplot(gse, showCategory = 1)

#pairwise_termsim(x, method = "JC", semData = NULL, showCategory = 200)

#gene <- names(gene_list)[abs(gene_list) > 2]
#ego <- enrichGO(gene = gene,
#universe = names(gene_list),
#OrgDb = org.Hs.eg.db,
#ont = "BP",
#pAdjustMethod = "BH",
#pvalueCutoff = 0.01,
#qvalueCutoff = 0.05,
#readable = TRUE)
#d <- godata('org.Hs.eg.db', ont="BP")
#ego2 <- pairwise_termsim(ego, method="Wang", semData = d)
#emapplot(ego2)
#emapplot_cluster(ego2)

```

##Category Netplot
The cnetplot depicts the linkages of genes and biological concepts (e.g. GO terms or KEGG pathways) as a network (helpful to see which genes are involved in enriched pathways and genes that may belong to multiple annotation categories).
```{r fig.width=18}

#install.packages("ggnewscale",dependencies = TRUE)
# categorySize can be either 'pvalue' or 'geneNum'
cnetplot(gseDvB, categorySize="pvalue", foldChange=gene_listDvB, showCategory = 3)
cnetplot(gseSexFvM, categorySize="pvalue", foldChange=gene_listSexFvM, showCategory = 3)
cnetplot(gseAgeOvY, categorySize="pvalue", foldChange=gene_listAgeOvY, showCategory = 3)

```

## Ridgeplot
Helpful to interpret up/down-regulated pathways.
```{r fig.width=18, fig.height=12}
install.packages("ggridges", dependencies = T)

ridgeplot(gse) + labs(x = "enrichment distribution")
ridgeplot(gse) + labs(x = "enrichment distribution") + theme(text = element_text(size=1)) + theme_minimal()

```

## GSEA Plot  
Traditional method for visualizing GSEA result.  
  
Params:  
**Gene Set** Integer. Corresponds to gene set in the gse object. The first gene set is 1, second gene set is 2, etc. 

```{r fig.height=6}
# Use the `Gene Set` param for the index in the title, and as the value for geneSetId
gseaplot(gse, by = "all", title = gse$Description[1], geneSetID = 1)
```

## PubMed trend of enriched terms
Plots the number/proportion of publications trend based on the query result from PubMed Central.
```{r fig.width=10}
install.packages("europepmc", dependencies = T)
terms <- gse$Description[1:3]
pmcplot(terms, 2010:2018, proportion=FALSE)
```

```{r}
library(DOSE)
# Convert gene IDs for gseKEGG function
# We will lose some genes here because not all IDs will be converted
idsDvB<-bitr(names(original_gene_listDvB), fromType = "SYMBOL", toType = "ENTREZID", OrgDb=organism)

idsSexFvM<-bitr(names(original_gene_listSexFvM), fromType = "SYMBOL", toType = "ENTREZID", OrgDb=organism)
idsAgeOvY<-bitr(names(original_gene_listAgeOvY), fromType = "SYMBOL", toType = "ENTREZID", OrgDb=organism)
idsC1<-bitr(names(original_gene_listC1), fromType = "SYMBOL", toType = "ENTREZID", OrgDb=organism)
idsC2<-bitr(names(original_gene_listC2), fromType = "SYMBOL", toType = "ENTREZID", OrgDb=organism)
idsC3<-bitr(names(original_gene_listC3), fromType = "SYMBOL", toType = "ENTREZID", OrgDb=organism)

# remove duplicate IDS (here I use "ENSEMBL", but it should be whatever was selected as keyType)
dedup_idsDvB = idsDvB[!duplicated(idsDvB[c("ENTREZID")]),]

dedup_idsSexFvM = idsSexFvM[!duplicated(idsSexFvM[c("ENTREZID")]),]
dedup_idsAgeOvY = idsAgeOvY[!duplicated(idsAgeOvY[c("ENTREZID")]),]
dedup_idsC1 = idsC1[!duplicated(idsC1[c("ENTREZID")]),]
dedup_idsC2 = idsC2[!duplicated(idsC2[c("ENTREZID")]),]
dedup_idsC3 = idsC3[!duplicated(idsC3[c("ENTREZID")]),]

# Create a new dataframe df2 which has only the genes which were successfully mapped using the bitr function above
DvB_uniqgenes2 = DvB_uniqgenes[DvB_uniqgenes$external_gene_name %in% dedup_idsDvB$SYMBOL,]
DvB_uniqgenes2$SYMBOL = DvB_uniqgenes2$external_gene_name
DvB_genesEntrez <- left_join(DvB_uniqgenes2,dedup_idsDvB, by="SYMBOL")

SexFvM_uniqgenes2 = SexFvM_uniqgenes[SexFvM_uniqgenes$external_gene_name %in% dedup_idsSexFvM$SYMBOL,]
SexFvM_uniqgenes2$SYMBOL = SexFvM_uniqgenes2$external_gene_name
SexFvM_genesEntrez <- left_join(SexFvM_uniqgenes2,dedup_idsSexFvM, by="SYMBOL")

AgeOvY_uniqgenes2 = AgeOvY_uniqgenes[AgeOvY_uniqgenes$external_gene_name %in% dedup_idsAgeOvY$SYMBOL,]
AgeOvY_uniqgenes2$SYMBOL = AgeOvY_uniqgenes2$external_gene_name
AgeOvY_genesEntrez <- left_join(AgeOvY_uniqgenes2,dedup_idsAgeOvY, by="SYMBOL")

C1_uniqgenes2 = Cluster1_uniqgenes[Cluster1_uniqgenes$external_gene_name %in% dedup_idsC1$SYMBOL,]
C1_uniqgenes2$SYMBOL = C1_uniqgenes2$external_gene_name
C1_genesEntrez <- left_join(C1_uniqgenes2,dedup_idsC1, by="SYMBOL")

C2_uniqgenes2 = Cluster2_uniqgenes[Cluster2_uniqgenes$external_gene_name %in% dedup_idsC2$SYMBOL,]
C2_uniqgenes2$SYMBOL = C2_uniqgenes2$external_gene_name
C2_genesEntrez <- left_join(C2_uniqgenes2,dedup_idsC2, by="SYMBOL")

C3_uniqgenes2 = Cluster3_uniqgenes[Cluster3_uniqgenes$external_gene_name %in% dedup_idsC3$SYMBOL,]
C3_uniqgenes2$SYMBOL = C3_uniqgenes2$external_gene_name
C3_genesEntrez <- left_join(C3_uniqgenes2,dedup_idsC3, by="SYMBOL")

# Create a new column in df2 with the corresponding ENTREZ IDs
#DvB_uniqgenes2$Y = dedup_idsDvB$ENTREZID

# Create a vector of the gene universe
kegg_gene_listDvB <- DvB_genesEntrez$log2FoldChange

kegg_gene_listSexFvM <- SexFvM_genesEntrez$log2FoldChange
kegg_gene_listAgeOvY <- AgeOvY_genesEntrez$log2FoldChange
kegg_gene_listC1 <- C1_genesEntrez$log2FoldChange
kegg_gene_listC2 <- C2_genesEntrez$log2FoldChange
kegg_gene_listC3 <- C3_genesEntrez$log2FoldChange

# Name vector with ENTREZ ids
names(kegg_gene_listDvB) <- DvB_genesEntrez$ENTREZID

names(kegg_gene_listSexFvM) <- SexFvM_genesEntrez$ENTREZID
names(kegg_gene_listAgeOvY) <- AgeOvY_genesEntrez$ENTREZID
names(kegg_gene_listC1) <- C1_genesEntrez$ENTREZID
names(kegg_gene_listC2) <- C2_genesEntrez$ENTREZID
names(kegg_gene_listC3) <- C3_genesEntrez$ENTREZID

# omit any NA values 
kegg_gene_listDvB<-na.omit(kegg_gene_listDvB)

kegg_gene_listSexFvM<-na.omit(kegg_gene_listSexFvM)
kegg_gene_listAgeOvY<-na.omit(kegg_gene_listAgeOvY)
kegg_gene_listC1<-na.omit(kegg_gene_listC1)
kegg_gene_listC2<-na.omit(kegg_gene_listC2)
kegg_gene_listC3<-na.omit(kegg_gene_listC3)

# sort the list in decreasing order (required for clusterProfiler)
kegg_gene_listDvB = sort(kegg_gene_listDvB, decreasing = TRUE)

kegg_gene_listSexFvM = sort(kegg_gene_listSexFvM, decreasing = TRUE)
kegg_gene_listAgeOvY = sort(kegg_gene_listAgeOvY, decreasing = TRUE)
kegg_gene_listC1 = sort(kegg_gene_listC1, decreasing = TRUE)
kegg_gene_listC2 = sort(kegg_gene_listC2, decreasing = TRUE)
kegg_gene_listC3 = sort(kegg_gene_listC3, decreasing = TRUE)

## Create gseKEGG object
 
**organism** KEGG Organism Code: The full list is here: https://www.genome.jp/kegg/catalog/org_list.html (need the 3 letter code). I define this as `kegg_organism` first, because it is used again below when making the pathview plots.  
**nPerm** permutation numbers, the higher the number of permutations you set, the more accurate your results is, but it will also cost longer time for running permutation.  
**minGSSize** minimal size of each geneSet for analyzing.   
**maxGSSize** maximal size of genes annotated for testing.   
**pvalueCutoff** pvalue Cutoff.   
**pAdjustMethod** one of "holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", "none".  
**keyType** one of 'kegg', 'ncbi-geneid', 'ncib-proteinid' or 'uniprot'.  

kegg_organism = "hsa"
kk2DvB <- gseKEGG(geneList     = kegg_gene_listDvB,
               organism     = kegg_organism,
               nPerm        = 10000,
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "none",
               keyType       = "ncbi-geneid")

kk2SexFvM <- gseKEGG(geneList     = kegg_gene_listSexFvM,
               organism     = kegg_organism,
               nPerm        = 10000,
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "none",
               keyType       = "ncbi-geneid")
kk2AgeOvY <- gseKEGG(geneList     = kegg_gene_listAgeOvY,
               organism     = kegg_organism,
               nPerm        = 10000,
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "none",
               keyType       = "ncbi-geneid")
kk2C1 <- gseKEGG(geneList     = kegg_gene_listC1,
               organism     = kegg_organism,
               nPerm        = 10000,
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "none",
               keyType       = "ncbi-geneid")
kk2C2 <- gseKEGG(geneList     = kegg_gene_listC2,
               organism     = kegg_organism,
               nPerm        = 10000,
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "none",
               keyType       = "ncbi-geneid")
kk2C3 <- gseKEGG(geneList     = kegg_gene_listC3,
               organism     = kegg_organism,
               nPerm        = 10000,
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "none",
               keyType       = "ncbi-geneid")

save(kk2DvB, file =   "kk2DvB.RData")
save(kk2SexFvM, file ="kk2SexFvM.RData")
save(kk2AgeOvY, file ="kk2AgeOvY.RData")
save(kk2C1, file =    "kk2C1.RData")
save(kk2C2, file =    "kk2C2.RData")
save(kk2C3, file =    "kk2C3.RData")

dotplot(kk2DvB, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)

dotplot(kk2SexFvM, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)
dotplot(kk2AgeOvY, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)
dotplot(kk2C1, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)
dotplot(kk2C2, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)
dotplot(kk2C3, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)

```
